Single-cell virus sequencing of influenza variants that trigger innate immunity

### 2018-09-05 paper/paper.tex
added computational methods
Added the computational methods section
as well as Supplementary files 1 and 2.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage[export]{adjustbox}
\DeclareSIUnit\Molar{M}

@@

@@

{\bf A} \includegraphics[width=0.9\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
\vspace{0.1in}

@@

\figsupp[Number of cells for which genotype(s) of infecting viruses were completely determined.]
{Cells for which we could determine the full sequences of all genes expressed by the infecting virus(es).
{\bf (A)} We could call the complete genotypes of the infecting virus(es) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNAs than cells for which we could not call complete genotypes.
This makes sense, as cells with more viral mRNA are more likely to have their viral genes captured in the PacBio sequencing, which is only able to capture a small fraction of the total transcripts identified by the 3'-sequencing of the 10X platform.
The lower calling rate for dual-barcode co-infections relative to single barcode infectionis is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing. 
An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
Note also that this plot is limited to the cells that were called as infected (\FIG{experiment}D) and could clearly be classified as IFN- or IFN+ (\FIG{experiment}H).
}
{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}}

@@

\subsection{Computational analysis of single-cell transcriptomic and viral sequence data}
A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}

@@

Add this

@@

>>>>

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage[export]{adjustbox}
\usepackage{float}
\DeclareSIUnit\Molar{M}

@@

\newfloat{suppfile}{thp}{lofsupfile}
%\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
\floatname{suppfile}{Supplementary file}

@@

\hspace{0.04in} {\bf A} \includegraphics[width=0.9\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
\vspace{0.1in}

@@

\figsupp[Number of cells for which genotype(s) of infecting viruses were completely determined.]
{Cells for which we could determine the full sequences of all genes expressed by the infecting virus(es).
{\bf (A)} We could call the complete genotypes of the infecting virus(es) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNAs than cells for which we could not call complete genotypes.
This makes sense, as cells with more viral mRNA are more likely to have their viral genes captured in the PacBio sequencing, which is only able to capture a small fraction of the total transcripts identified by the 3'-sequencing of the 10X platform.
The lower calling rate for dual-barcode co-infections relative to single barcode infectionis is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing. 
An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
Note also that this plot is limited to the cells that were called as infected (\FIG{experiment}D) and could clearly be classified as IFN- or IFN+ (\FIG{experiment}H).
}
{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}}
\label{figsupp:ncells}

@@

\subsection{Computational analysis of single-cell transcriptomic and viral sequence data}
A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}~\citep{russell2018github}. 
This pipeline is orchestrated by \texttt{Snakemake}~\citep{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
The sequencing data and annotated cell-gene matrix are available on the GEO repository under \jdbcomment{add accession}.

Briefly, the raw deep sequencing data from the Illumina 3'-end sequencing were processed using the 10X Genomics software package \texttt{cellranger} (version 2.2.0). 
We built a multi-species alignment reference consisting of a concatenation of the human and influenza virus transcriptomes (the first ``species'') and the canine transcriptome (the second ``species''). 
The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
The influenza virus transcriptome consisted of the mRNAs for the wildtype A/WSN/1933 virus strain in \FIGDATA[experiment]{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wildtype and synonymously barcoded viral variants to this transcriptome).
The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
The \texttt{cellranger} software was used to align the Illumina 3'-end sequencing reads to this multi-species transcriptome, call human+influenza and canine cells (\FIG{experiment}B), and generate a matrix giving the expression of each gene in each single cell.
We used a custom Python script to determine the number of influenza virus reads that could be assigned to the wildtype or synonymously barcoded virus, and added this information to the annotated the cell-gene matrix.

The PacBio sequences of the full-length viral genes were analyzed as follows.
First, we used version 3.1.0 of PacBio's \texttt{ccs} program (\url{https://github.com/PacificBiosciences/unanimity}) to build circular consensus sequences (CCSs) from the subreads files, requiring at least 3 passes and a minimum accuracy of 0.999.
We further processed these CCSs using custom Python code and the \texttt{minimap2}~\citep{li2018minimap2} long-read aligner (version 2.11-r797).
The Python code has been implemented in the API of \texttt{dms\_tools2}~\citep[][\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software} package (version 2.3.0).
A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as Supplementary~file~\ref{suppfile:pacbio_analysis}.
We refer the reader to this notebook for a detailed description and extensive plots showing the results at each step.
Here is a brief summary: we filtered for CCSs that had the expected 5' termini (from the influenza-specific primers) and 3' termini (corresponding to the 10X adaptor), and for which we could identify the cell barcode, UMI, and polyA tail.
We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\FIGSUPP[genotypes]{StrandExchange}), and reassuringly found that strand exchange was rare (\FIGSUPP[genotypes]{CCSs}).
The small number of CCSs with identifiable strand exchange were filtered from further analysis.
We then further filtered for CCSs that contained valid cell barcodes as identified by the \texttt{cellranger} pipeline, and kept just one CCS per UMI (preferentially retaining high-quality CCSs that aligned to full-length cDNAs).
Finally, we used the CCSs to call the sequence of the viral gene in each cell, calling mutations separately for each viral barcode variant.
We called mutations (insertions, deletions, and substitutions) in the viral gene sequences as follows:
\begin{enumerate}
\item Mutations with accuracies less than 0.999 (which constitute $<$0.5\% of all mutations) were ignored.
\item If all CCSs for a particular viral-barcode variant of a gene in a cell were wildtype, it was called as wildtype.
\item If any CCSs for a particular viral-barcode variant of gene in a cell had a mutation, then require at least two CCSs to call the sequence.
\item If at least two and $>$30\% of the CCSs had a specific mutation, then call that mutation as present and note its frequency among the CCSs. The exception was single-nucleotide indels in homopolymers, for which we required three CCSs to call a mutation (the reason is that the main mode of PacBio sequencing errors is short indels in homopolymers).
\end{enumerate}
The plots in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or Supplementary~file~\ref{suppfile:pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
We could call the sequences of all expressed viral genes in about half of the infected cells (\FIGSUPP[genotypes]{ncells}).
The mutations called using this pipeline are shown in \FIG{genotypes}, and \FIGDATA[genotypes]{genotypes} gives the number of CCSs supporting each mutation call.
The called sequences of the viral genes were added to the annotated cell-gene matrix.

Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
This analysis utilized a variety of R and Bioconductor~\citep{huber2015orchestrating} packages, including \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as Supplementary~file~\ref{suppfile:monocle_analysis}.
We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in \FIG{experiment}C.
We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{experiment}D).
We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{experiment}E and \FIGSUPP[experiment]{frac_has_gene}). 
To identify cells that were co-infected with both viral barcodes (\FIG{experiment}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived form the more common viral barcode.
We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{experiment}H and \FIGSUPP[experiment]{ISG}, respectively.
We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
The plot in \FIG{genotypes} summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (\url{https://github.com/wilkox/gggenes}).

@@

\jdbcomment{Add this}

@@

\clearpage

\begin{suppfile}
\caption{\label{suppfile:pacbio_analysis}
An HTML rendering of the Jupyter notebook that analyzes the PacBio data to call the viral sequences in infected cells is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/pacbio_single_cell_figures/pacbio_analysis.html}.
This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb.}}
\end{suppfile}

\begin{suppfile}
\caption{\label{suppfile:monocle_analysis}
An HTML rendering of the Jupyter notebook that analyzes the annotated cell-gene matrix to generate the figures in this paper is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb.}}
\end{suppfile}
